Reinforcement Learning Produces Dominant Strategies for the Iterated Prisoner's Dilemma

### 2017-07-14 main.tex
Update minor things.
- Fix percentage not appearing
- Tweak the abstract
- Consistant IPD
- Include some more references.

\begin{abstract}
    We present tournament results and several powerful strategies for the Iterated
    Prisoner's Dilemma created using reinforcement learning techniques
    (evolutionary and particle swarm algorithms). These strategies are
    trained to perform well against a corpus of over 170 distinct
    opponents, including many well-known and classic strategies. All
    the trained strategies win standard tournaments against the total collection
    of other opponents. Variants are also trained to win noisy tournaments.
\end{abstract}

@@

The iterated prisoner's dilemma (IPD) is a common model in game theory,
frequently used  to understand the evolution of cooperative behaviour from complex
dynamics \cite{Axelrod1984}.

In this manuscript we use the Axelrod library \cite{axelrodproject, knight2016open},
open source software for conducting IPD research with
reproducibility as a principal goal. Written in the Python programming language,
to date the library contains source code contributed by over 50 individuals
from a variety of geographic locations and technical backgrounds. The library
is supported by a comprehensive test suite that covers all the intended
behaviors of all of the strategies in the library, as well as the features that
conduct matches, tournaments, and population dynamics.

@@

In this work we describe how collections of strategies in the Axelrod library
have been used to train new strategies specifically to win IPD tournaments.
These strategies are trained using generic strategy archetypes based on e.g.\
finite state
machines, arriving at particularly effective parameter choices through
evolutionary or particle swarm algorithms. There are several
previous publications that use evolutionary algorithms to
evolve IPD strategies in various circumstances
\cite{ashlock2006training, Ashlock2015, Ashlock2006,
      ashlock2014shaped, Ashlock2014, barlow2015varying,
      fogel1993evolving, marks1989niche, sudo2015effects,
      vassiliades2010multiagent}. See also \cite{Gaudesi2016} for a
strategy trained to win against a collection of well-known IPD opponents and see
\cite{franken2005particle} for a prior use of particle swarm algorithms. Our
results are unique in that we are able to train against a large and diverse
collection of strategies available from the scientific literature.
Crucially, the
software used in this work is openly available and can be used to train strategies
in the future in a reliable manner, with confidence that the opponent strategies
are correctly implemented, tested and documented.
Moreover, as of the time of writing, we claim that this work contains the best
performing strategies for the iterated prisoner's dilemma.

@@

The Axelrod library now contains many parametrised strategies trained using
machine learning
methods. Most are deterministic, use many rounds of memory, and perform
extremely well in tournaments as will be discussed in Section~\ref{sec:results}.
Training of these strategies will be discussed in Section~\ref{sec:methods}.
These strategies can encode a variety
of other strategies, including classic strategies like Tit For Tat,
handshake strategies, and grudging strategies that always defect after
an opponent defection.

@@

This section presents the results of a large IPD tournament with
strategies from the Axelrod library, including some additional parametrized
strategies (e.g.\ various parameter choices for Generous Tit For Tat). These are
listed in Appendix~\ref{app:list_of_players}.

@@

The top 11 performing strategies by median payoff are all strategies trained to maximize
total payoff against a subset of the strategies (Table~\ref{tbl:standard_score}).
The next strategy is Desired Belief Strategy (DBS) \cite{Au2006}.
which actively analyzes the opponent and responds
accordingly. The next two strategies are Winner12, based on a lookup table,
Fool Me Once, a grudging strategy that defects indefinitely on
the second defection, and Omega Tit For Tat \cite{kendall2007iterated}.

@@

Finally, Table~\ref{tbl:noisy_ranks} and
Figure~\ref{fig:noisy_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. We see that the stochasticity
of the ranks understandably increases relative to the standard tournament. An
exception is the top three strategies, for example, the DBS strategy never ranks
lower than
second and wins 75\% of the time. The two strategies trained for noisy
tournaments rank in the top three 95\% of the time.
% TODO Possibly update this with more data.

@@

In opposition to historical tournament results and community folklore,
our results show that complex strategies can be very effective for the
IPD\@. Designing complex strategies for the prisoner's dilemma appears to be
difficult for humans. Of all the human-designed
strategies in the library, only DBS consistently performs well, and it is
substantially more complex than traditional tournament winners like TFT, OmegaTFT,
and zero determinant strategies. Furthermore, dealing with noise is difficult
for most strategies. Two strategies designed specifically to account for noise,
DBS and OmegaTFT, perform well and only DBS performs better than the trained
strategies and \textbf{only} in some noisy contexts. Empirically we find that
DBS (with its default parameters) does not win tournaments at 1% noise.
However DBS has a parameter that accounts for the expected amount of noise and a
followup study with various noise levels could make a more complete study of
the performance of DBS and strategies trained at various noise levels.

>>>>

\begin{abstract}
    We present tournament results and several powerful strategies for the Iterated
    Prisoner's Dilemma created using reinforcement learning techniques
    (evolutionary and particle swarm algorithms). These strategies are
    trained to perform well against a corpus of over 170 distinct
    opponents, including many well-known and classic strategies. All
    the trained strategies win standard tournaments against the total collection
    of other opponents. The trained strategies and one particular human made
    designed strategy are the top performers in noisy tournaments also.
\end{abstract}

@@

The Iterated Prisoner's Dilemma (IPD) is a common model in game theory,
frequently used  to understand the evolution of cooperative behaviour from complex
dynamics \cite{Axelrod1984}.

In this manuscript the Axelrod library \cite{knight2016open, axelrodproject},
open source software for conducting IPD research with
reproducibility as a principal goal, is used. Written in the Python programming language,
to date the library contains source code contributed by over 50 individuals
from a variety of geographic locations and technical backgrounds. The library
is supported by a comprehensive test suite that covers all the intended
behaviors of all of the strategies in the library, as well as the features that
conduct matches, tournaments, and population dynamics.

@@

In this work we describe how collections of strategies in the Axelrod library
have been used to train new strategies specifically to win IPD tournaments.
These strategies are trained using generic strategy archetypes based on e.g.\
finite state
machines, arriving at particularly effective parameter choices through
evolutionary or particle swarm algorithms. There are several
previous publications that use evolutionary algorithms to
evolve IPD strategies in various circumstances
\cite{ashlock2006training, Ashlock2015, Ashlock2006,
      ashlock2014shaped, Ashlock2014, barlow2015varying,
      fogel1993evolving, marks1989niche, sudo2015effects,
      vassiliades2010multiagent}. See also \cite{Gaudesi2016} for a
strategy trained to win against a collection of well-known IPD opponents and see
\cite{franken2005particle} for a prior use of particle swarm algorithms. Our
results are unique in that we are able to train against a large and diverse
collection of strategies available from the scientific literature.
Crucially, the
software used in this work is openly available and can be used to train strategies
in the future in a reliable manner, with confidence that the opponent strategies
are correctly implemented, tested and documented.
Moreover, as of the time of writing, we claim that this work contains the best
performing strategies for the Iterated Prisoner's Dilemma.

@@

The Axelrod library now contains many parametrised strategies trained using
machine learning
methods. Most are deterministic, use many rounds of memory, and perform
extremely well in tournaments as will be discussed in Section~\ref{sec:results}.
Training of these strategies will be discussed in Section~\ref{sec:methods}.
These strategies can encode a variety
of other strategies, including classic strategies like Tit For Tat
\cite{Axelrod1980},
handshake strategies, and grudging strategies that always defect after
an opponent defection.

@@

This section presents the results of a large IPD tournament with
strategies from the Axelrod library, including some additional parametrized
strategies (e.g.\ various parameter choices for Generous Tit For Tat
\cite{Gaudesi2016}). These are
listed in Appendix~\ref{app:list_of_players}.

@@

The top 11 performing strategies by median payoff are all strategies trained to maximize
total payoff against a subset of the strategies (Table~\ref{tbl:standard_score}).
The next strategy is Desired Belief Strategy (DBS) \cite{Au2006}.
which actively analyzes the opponent and responds
accordingly. The next two strategies are Winner12, based on a lookup table,
Fool Me Once \cite{axelrodproject}, a grudging strategy that defects indefinitely on
the second defection, and Omega Tit For Tat \cite{kendall2007iterated}.

@@

Finally, Table~\ref{tbl:noisy_ranks} and
Figure~\ref{fig:noisy_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. We see that the stochasticity
of the ranks understandably increases relative to the standard tournament. An
exception is the top three strategies, for example, the DBS strategy never ranks
lower than
second and wins 75\% of the time. The two strategies trained for noisy
tournaments rank in the top three 95\% of the time.

@@

In opposition to historical tournament results and community folklore,
our results show that complex strategies can be very effective for the
IPD\@. Designing complex strategies for the prisoner's dilemma appears to be
difficult for humans. Of all the human-designed
strategies in the library, only DBS consistently performs well, and it is
substantially more complex than traditional tournament winners like TFT, OmegaTFT,
and zero determinant strategies. Furthermore, dealing with noise is difficult
for most strategies. Two strategies designed specifically to account for noise,
DBS and OmegaTFT, perform well and only DBS performs better than the trained
strategies and \textbf{only} in some noisy contexts. Empirically we find that
DBS (with its default parameters) does not win tournaments at 1\% noise.
However DBS has a parameter that accounts for the expected amount of noise and a
followup study with various noise levels could make a more complete study of
the performance of DBS and strategies trained at various noise levels.
